experDesign: stratifying samples into batches with minimal bias

### 2021-11-08 paper.Rmd
Remove long lines moves tatement of need below code examples
Update paper

---
title: 'experDesign: stratifying samples into batches with minimal bias'
output:
  word_document:
    keep_md: yes
  md_document:
    preserve_yaml: yes
  pdf_document: default
  html_document:
    df_print: paged
date: "25th August 2021"
affiliations:
- name: Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas
  index: 1
- name: Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS
  index: 2
authors:
- name: Lluís Revilla Sancho
  orcid: 0000-0001-9747-2570
  affiliation: 1, 2
- name: Juan-José Lozano
  orcid: 0000-0001-9747-2570
  affiliation: '1'
- name: Azucena Salas
  orcid: 0000-0001-9747-2570
  affiliation: '2'
tags:
- R
- batch effect
- experiment design
bibliography: paper.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, warning = TRUE, error = TRUE, fig.width = 9)
```

@@

@@

For completeness a description and comparison of the usage of the different software packages currently available on CRAN and Bioconductor is presented below.
First, we start with some real data obtained from a survey:

```{r example-data}
data(survey, package = "MASS")
survey$ID <- seq_len(nrow(survey))
VoI <- c("Sex", "Smoke", "Age")
n_batch <- 3
size_subset <- 96
iterations <- 1000
```

This data set has three variables of interest; Sex, Smoke and Age are a mix of categorical and numeric variables.

```{r osat, fig.cap="Sample distributions with OSAT."}
library("ggplot2")
theme_plots <- theme_minimal() + theme(text = element_text(size = 10))
theme_set(theme_plots)
library("OSAT")
library("patchwork")
gs <- setup.sample(survey, optimal = VoI)
gc <- setup.container(plate = IlluminaBeadChip96Plate, n = n_batch, batch = 'plates')
gSetup <- create.optimized.setup(sample = gs, container = gc, nSim = iterations)
osat_report <- get.experiment.setup(gSetup)
nrow(osat_report)
osat_age <- ggplot(osat_report) +
  geom_histogram(aes(Age, fill = as.character(plates))) +
  facet_wrap(~plates) +
  labs(fill = "Batch")
osat_sex <- ggplot(osat_report) +
  geom_bar(aes(Sex, fill = as.character(plates))) +
  facet_wrap(~plates) +
  labs(fill = "Batch")
osat_smoke <- ggplot(osat_report) +
  geom_bar(aes(Smoke, fill = as.character(plates))) +
  facet_wrap(~plates) +
  labs(fill = "Batch")
(osat_smoke/(osat_age + osat_sex + plot_layout(guides = 'collect'))) +
  plot_annotation(title = "OSAT")
```

OSAT returned one row less than the input provided.
This row was deleted because it had an NA value for the Sex variable.

anticlust does not handle all the variables, it only accepts numeric variables:

```{r anticlust, error=TRUE, fig.cap="Sample distributions with anticlust."}
library("anticlust")
anticlust_index <- anticlustering(
  survey[, VoI],
  K = n_batch,
  objective = "variance",
  method = "exchange",
  repetitions = iterations
)
anticlust_index <- anticlustering(
  survey[, "Age"],
  K = n_batch,
  objective = "variance",
  method = "exchange",
  repetitions = iterations
)
report_anticlust <- data.frame(Age = survey[, "Age"], batch = as.character(anticlust_index))
library("ggplot2")
ggplot(report_anticlust) + 
  geom_histogram(aes(Age, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch", title = "anticlust")
```

There is a bug in the Omixer that prevents it from working unless specific conditions are meet.
This precluded any comparisons of Omixer with other tools using the same settings:

```{r Omixer, error=TRUE}
library("Omixer")
survey$ID <- seq_len(nrow(survey))
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, 
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(1, nrow(survey)),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(0, nrow(survey)),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(1, 288),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(0, 288),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)

@@

If users want a design without replicates but the batches have some spatial distribution, we must use `design` to allocate the samples on each batch, followed by the `spatial` function to randomly distribute the samples homogeneously by position within each batch.
See the example below:

```{r experDesign, fig.cap="Sample distributions with experDesign."}
library("experDesign")
experDesign_index <- design(survey[, VoI], size_subset = size_subset,
                  iterations = iterations, name = "")
experDesign_index_spatial <- spatial(index = experDesign_index, pheno = survey[, VoI], 
                                     iterations = iterations, rows = LETTERS[1:8], columns = 1:12)
experDesign_report <- inspect(experDesign_index, survey[, VoI])
nrow(experDesign_report)
experDesign_age <- ggplot(experDesign_report) + 
  geom_histogram(aes(Age, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
experDesign_sex <- ggplot(experDesign_report) +
  geom_bar(aes(Sex, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
experDesign_smoke <- ggplot(experDesign_report) +
  geom_bar(aes(Smoke, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
(experDesign_smoke/(experDesign_age + experDesign_sex + plot_layout(guides = 'collect'))) + plot_annotation(title = "experDesign")
```

The statistics of the index can be checked for multiple statistics, as shown in the following code:

```{r evaluate}
evaluate_na(experDesign_index, survey[, VoI])
evaluate_entropy(experDesign_index, survey[, VoI])
evaluate_mad(experDesign_index, survey[, VoI])
evaluate_sd(experDesign_index, survey[, VoI])
evaluate_mean(experDesign_index, survey[, VoI])
# All together for each batch
ei <- evaluate_index(experDesign_index, survey[, VoI])
ei[, , "SubSet1"]
evaluate_na(experDesign_index_spatial, survey[, VoI])
evaluate_entropy(experDesign_index_spatial, survey[, VoI])
evaluate_mad(experDesign_index_spatial, survey[, VoI])
evaluate_sd(experDesign_index_spatial, survey[, VoI])
evaluate_mean(experDesign_index_spatial, survey[, VoI])
# All together for each batch
ei <- evaluate_index(experDesign_index_spatial, survey[, VoI])
ei[, , "G1"]
```

We can also compare our results with the original distribution:

```{r original}
evaluate_orig(survey[, VoI])
```

In addition to distributing the samples into batches, ***experDesign*** provides tools to add technical replicates.
In order to choose them from the available samples, the function `extreme_cases` is provided.
For easier usage, the `replicates` function designs an experiment with the desired number of replicates per batch.
The rest of the workflow is similar to the above example:

```{r controls}
extreme_cases(survey[, VoI], size = 5)
replicates_index <- replicates(pheno = survey[, VoI], 
                               size_subset = size_subset, controls = 5, 
                               iterations = iterations)
spatial_replicate_index <- spatial(index = replicates_index, pheno = survey[, VoI], 
                                   rows = LETTERS[1:8], columns = 1:12)
survey$ID <- seq_len(nrow(survey))
report_replicates <- inspect(i = replicates_index, 
                             pheno = survey[, c(VoI, "ID")])
report_replicates_position <- inspect(i = spatial_replicate_index, 
                             pheno = report_replicates, 
                             index_name = "position")
# Batch and position where the samples should be:
head(report_replicates_position)
# Technical replicates:
head(report_replicates_position[duplicated(report_replicates_position$ID), ])
```

@@

```{r show-batches}
optimum_batches(size_data = 250, size_subset = 96)
sizes_batches(size_data = 250, size_subset = 96, batches = 3)
```

>>>>

---
title: 'experDesign: stratifying samples into batches with minimal bias'
output:
  html_document:
    df_print: paged
  word_document:
    keep_md: yes
  pdf_document: default
  md_document:
    preserve_yaml: yes
date: "25th August 2021"
affiliations:
- name: Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas
  index: 1
- name: Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS
  index: 2
authors:
- name: Lluís Revilla Sancho
  orcid: 0000-0001-9747-2570
  affiliation: 1, 2
- name: Juan-José Lozano
  orcid: 0000-0001-9747-2570
  affiliation: '1'
- name: Azucena Salas
  orcid: 0000-0001-9747-2570
  affiliation: '2'
tags:
- R
- batch effect
- experiment design
bibliography: paper.bib
---

```{r setup}
knitr::opts_chunk$set(collapse = TRUE, warning = TRUE, 
                      error = 0,
                      fig.width = 9)
```

@@

For completeness a description and comparison of the usage of the different software packages currently available on CRAN and Bioconductor is presented below.
First, we start with some real data obtained from a survey.
This data set has three variables of interest; Sex, Smoke and Age are a mix of categorical and numeric variables.

```{r example-data, include = FALSE}
data(survey, package = "MASS")
survey$ID <- seq_len(nrow(survey))
VoI <- c("Sex", "Smoke", "Age")
n_batch <- 3
size_subset <- 96
iterations <- 1000
```

```{r osat, fig.cap="Sample, distributions with OSAT.", include = FALSE}
library("ggplot2")
theme_plots <- theme_minimal() + theme(text = element_text(size = 10))
theme_set(theme_plots)
library("OSAT")
library("patchwork")
gs <- setup.sample(survey, optimal = VoI)
gc <- setup.container(plate = IlluminaBeadChip96Plate, 
                      n = n_batch, batch = 'plates')
gSetup <- create.optimized.setup(sample = gs, container = gc, 
                                 nSim = iterations)
osat_report <- get.experiment.setup(gSetup)
nrow(osat_report)
osat_age <- ggplot(osat_report) +
  geom_histogram(aes(Age, fill = as.character(plates))) +
  facet_wrap(~plates) +
  labs(fill = "Batch")
osat_sex <- ggplot(osat_report) +
  geom_bar(aes(Sex, fill = as.character(plates))) +
  facet_wrap(~plates) +
  labs(fill = "Batch")
osat_smoke <- ggplot(osat_report) +
  geom_bar(aes(Smoke, fill = as.character(plates))) +
  facet_wrap(~plates) +
  labs(fill = "Batch")
```

OSAT might return less rows than the input provided because they might have NA value.

anticlust does not handle all the variables, it only accepts numeric variables.

```{r anticlust, fig.cap="Sample, distributions with anticlust.", include = FALSE}
library("anticlust")
anticlust_index <- anticlustering(
  survey[, VoI],
  K = n_batch,
  objective = "variance",
  method = "exchange",
  repetitions = iterations
)
anticlust_index <- anticlustering(
  survey[, "Age"],
  K = n_batch,
  objective = "variance",
  method = "exchange",
  repetitions = iterations
)
report_anticlust <- data.frame(Age = survey[, "Age"], 
                               batch = as.character(anticlust_index))
library("ggplot2")
anticlust <- ggplot(report_anticlust) + 
  geom_histogram(aes(Age, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
```

There is a bug in the Omixer that prevents it from working unless specific conditions are meet.
This precluded any comparisons of Omixer with other tools using the same settings.

```{r Omixer, include = FALSE}
library("Omixer")
survey$ID <- seq_len(nrow(survey))
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, 
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, 
                           mask = rep(1, nrow(survey)),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, 
                           mask = rep(0, nrow(survey)),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, 
                           mask = rep(1, 288),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, 
                           mask = rep(0, 288),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)

```{r panel1, echo=FALSE, message=FALSE, warning=FALSE}
(osat_smoke/(osat_age + osat_sex) / anticlust & plot_layout(guides = 'collect')
```

@@

@@

If users want a design without replicates but the batches have some spatial distribution, we must use `design` to allocate the samples on each batch, followed by the `spatial` function to randomly distribute the samples homogeneously by position within each batch.
See the example in `inspect` and the vignette:

```{r experDesign, echo=FALSE, fig.cap="Sample, distributions with anticlust.", warning=FALSE, message=FALSE}
library("experDesign")
experDesign_index <- design(survey[, VoI], size_subset = size_subset,
                  iterations = iterations, name = "")
experDesign_index_spatial <- spatial(index = experDesign_index, 
                                     pheno = survey[, VoI], 
                                     iterations = iterations, 
                                     rows = LETTERS[1:8], 
                                     columns = 1:12)
experDesign_report <- inspect(experDesign_index, survey[, VoI])
nrow(experDesign_report)
experDesign_age <- ggplot(experDesign_report) + 
  geom_histogram(aes(Age, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
experDesign_sex <- ggplot(experDesign_report) +
  geom_bar(aes(Sex, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
experDesign_smoke <- ggplot(experDesign_report) +
  geom_bar(aes(Smoke, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
(experDesign_smoke/(experDesign_age + 
                      experDesign_sex )) +
  plot_annotation(title = "experDesign") +
  plot_layout(guides = 'collect')
```

The statistics of the index can be checked for multiple statistics, as shown on the help pages of `evaluate_na`, `evaluate_entropy`, `evaluate_mad`, `evaluate_sd` and `evaluate_mean`.

```{r evaluate, include = FALSE}
evaluate_na(experDesign_index, survey[, VoI])
evaluate_entropy(experDesign_index, survey[, VoI])
evaluate_mad(experDesign_index, survey[, VoI])
evaluate_sd(experDesign_index, survey[, VoI])
evaluate_mean(experDesign_index, survey[, VoI])
# All together for each batch
ei <- evaluate_index(experDesign_index, survey[, VoI])
ei[, , "SubSet1"]
evaluate_na(experDesign_index_spatial, survey[, VoI])
evaluate_entropy(experDesign_index_spatial, survey[, VoI])
evaluate_mad(experDesign_index_spatial, survey[, VoI])
evaluate_sd(experDesign_index_spatial, survey[, VoI])
evaluate_mean(experDesign_index_spatial, survey[, VoI])
# All together for each batch
ei <- evaluate_index(experDesign_index_spatial, survey[, VoI])
ei[, , "G1"]
```

We can also compare our results with the original distribution via `evaluate_orig`.

```{r original, include = FALSE}
evaluate_orig(survey[, VoI])
```

In addition to distributing the samples into batches, ***experDesign*** provides tools to add technical replicates.
In order to choose them from the available samples, the function `extreme_cases` is provided.
For easier usage, the `replicates` function designs an experiment with the desired number of replicates per batch.

```{r controls, include = FALSE}
extreme_cases(survey[, VoI], size = 5)
replicates_index <- replicates(pheno = survey[, VoI], 
                               size_subset = size_subset, controls = 5, 
                               iterations = iterations)
spatial_replicate_index <- spatial(index = replicates_index, 
                                   pheno = survey[, VoI], 
                                   rows = LETTERS[1:8], columns = 1:12)
survey$ID <- seq_len(nrow(survey))
report_replicates <- inspect(i = replicates_index, 
                             pheno = survey[, c(VoI, "ID")])
report_replicates_position <- inspect(i = spatial_replicate_index, 
                             pheno = report_replicates, 
                             index_name = "position")
# Batch and position where the samples should be:
head(report_replicates_position)
# Technical replicates:
head(report_replicates_position[duplicated(report_replicates_position$ID), ])
```

@@

```{r show-batches, include = FALSE}
optimum_batches(size_data = 250, size_subset = 96)
sizes_batches(size_data = 250, size_subset = 96, batches = 3)
```
